---
title: "Creating Summary Lines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Summary Lines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options, message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")

library(gt)
library(dplyr)
```

Tables with summary data are quite common and naturally occur with alarming regularity. We can use **gt** to produce such summary components. These summary rows are automatically inserted within the relevant stub blocks, where each block is treated as a group and where each summary row is the result of a different aggregation function. 

### Preparing the input data table

To most easily insert the necessary summary rows, we need to ensure that some preparatory work is done. The key change to the input data table should be to have a `groupname` column present, with categorical group names for each grouping. The following example data table contains both `groupname` and `rowname` columns, along with two data columns (named `value_1` and `value_2`). Having the magic column names `groupname` and `rowname` will signal to **gt** that they should be used to form a table stub where `groupname` collects rows into a stub block (furnishing a group heading row above), and `rowname` provides row labels within each stub block. Here is a diagram showing the restructuring:

The incoming data table for the purpose of the first example will be created using `dplyr::tribble()`:

```{r}
# Create a table that creates a stub and stub blocks by
# virtue of having `groupname` and `rowname` columns
tbl <-
  dplyr::tribble(
    ~groupname, ~rowname, ~value_1, ~value_2,
    "A",        "1",      235.6,    260.1,
    "A",        "2",      184.3,    84.4,
    "A",        "3",      342.3,    126.3,
    "A",        "4",      234.9,    83.0,
    "B",        "1",      190.9,    832.5,
    "B",        "2",      743.3,    281.2,
    "B",        "3",      252.3,    732.5,
    "B",        "4",      344.7,    546.1,
    "C",        "1",      197.2,    818.0,
    "C",        "2",      284.3,    394.4)
```

There are three groups in this data table: `A`, `B`, and `C` and summaries will be made within specified groups using the `summary_rows()` function. There’s a lot of control in how the summaries are conducted and formatted. First of all, we can choose which groups will receive summary rows (the rows appear at the bottom of each stub block). This optional choice of groups can be made using the `groups` argument. Secondly, we can choose which columns should be included in the summary with the `columns` argument. For both arguments, we can use the `vars()` helper function to easily specify the groups and columns using bare names.

To make any sort of summary, we need to use functions that will perform the aggregation. We can provide base functions such as `mean()`, `sum()`, `min()`, `max()`, and more, using the `funs` argument. It’s preferable here to use the `funs()` helper function to supply a list of aggregation functions. Each function provided will result in a summary row for each group.

Because each function will yield a row, we need to be able to identify that row. So, each summary row will receive a row caption. We can provide our preferred names as a character vector to the `labels` argument, in the same order and quantity as in `funs`. If nothing is supplied, then **gt** will use the input with the `funs` argument to generate the row captions.

Now that the `summary_rows()` function has been somewhat explained, let’s look at how we can get summary rows for the `tbl` table:

```{r}
# Create a gt table with summary rows
gt_summary_a <-
  gt(tbl) %>%
  summary_rows(
    groups = vars(A, B),
    columns = vars(value_1),
    funs = funs(mean, sum, sd))
```

```{r}
# View the HTML table
gt_summary_a
```

Here we see that summary rows were created for the `A` and `B` groups (group `C` has no summary lines in its stub block). It was specified (in `columns`) that only the `value_1` column should be evaluated and, indeed, the `value_2` column shows no values within the summary rows.

There is some flexibility in how we supply functions to the `funs()` helper function. We can also use function names enclosed in quotes (e.g., `funs("mean", "sum", "sd")`) or a vector of function names (e.g., `funs(c("mean", "sum", "sd))`). This is nice if we want to build a vector of function names and perhaps pass in the same vector to `labels`. Here's an example that creates the extact same output table as before (so we won't show the result):

```{r eval=FALSE, include=TRUE}
# Define a vector of function names
function_vec <- c("mean", "sum", "sd")

# Use `function_vec` in `funs` and `labels`
gt_summary_b <-
  gt(tbl) %>%
  summary_rows(
    groups = vars(A, B),
    columns = vars(value_1),
    funs = funs(function_vec),
    labels = c("average", "total", "std dev"))
```

### Formatting the summary output values

The `summary_rows()` function provides the following arguments for formatting the output values:

- `decimals`: the exact number of decimal places to use (default is `2`)
- `sep_mark`: the mark to use as a separator between groups of digits (default is `,`)
- `dec_mark`: the character to use as a decimal mark (default is `.`)

These options are similar to what is available in the `fmt_number()` function. We can base the formatting of the summary row values to those values that supply inputs to the summary rows. Here is an example that uses the formatting options in `fmt_number()` and in `summary_rows()` to provide numeric values that are formatted to 3 decimal places throughout the entire table:

```{r}
# Format the numeric row data to have 3 decimal
# place and do the same for the summary row values
gt_summary_c <-
  gt(tbl) %>%
  fmt_number(
    columns = vars(value_1, value_2),
    decimals = 3) %>%
  summary_rows(
    groups = vars(A, B),
    columns = vars(value_1, value_2),
    funs = funs(min, max, mean),
    labels = c("minimum", "maximum", "average"),
    decimals = 3)

gt_summary_c
```

### Extracting the summary data from the gt table object

For a reproducible workflow, we do not want to have situtations where any data created or modified cannot be accessed. While having summarized values be created in a **gt** pipeline presents advantages to readability and intent of analysis, it is recognized that the output table itself is essentially 'read only', as the input data undergoes processing and movement to an entirely different format.

However, the object created still contains data and we can obtain the summary data from a gt table object using the `extract_summary()` function. Taking the `gt_summary` object, we get a data frame containing the summary data while preserving the correct data types:

```{r}
gt_summary_c %>% extract_summary()
```

The output data frame contains the `groupname` and `rowname` columns. The `groupname` column presents the stub groups that were given in `summary_rows()`.  The `rowname` column contains the descriptive stub labels for the summary rows (recall that values are either supplied explicitly in `summary_rows()`, or, are generated). The remaining columns are a subset of the original dataset, where only those columns that had aggregation functions applied are retained.

The output data frame from `extract_summary()` can be reintroduced to a reproducible workflow and serve as downstream inputs or undergo validation. Perhaps interestingly, the output data frame is structured in such a way that it is suitable for direct generation of a **gt** table with `gt()` (serving as a standalone summary table):

```{r}
gt_summary_c %>%
  extract_summary() %>%
  gt() %>%
  fmt_number(
    columns = vars(value_1, value_2),
    decimals = 3)
```

