---
title: "gt interactive tables"
format: html
html-table-processing: none
description: >
  An overview of interactive tables with gt
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
set.seed(123)
```

gt provides an option to make interactive html tables via the [reactable](https://glin.github.io/reactable/index.html) package.

It also provides an interactive to creating gt tables as plots.

Let's use the following base for our gt table.

```{r}
 devtools::load_all("~/rrr-forks/gt")
#library(gt)
gt_tbl <- exibble |>
  gt(groupname_col = "group", rowname_col = "row") |>
  tab_header(
    "Title",
    "Subtitle"
  ) |>
  tab_footnote(
    "A footnote"
  ) |>
  tab_spanner(
    "Spanner",
    columns = c(date, time)
  )
```

To create an interactive table, you have simply have to pipe `opt_interactive()` to your existing gt pipeline.

::: {.panel-tabset}

## Html

```{r}
gt_tbl
```

## Interactive

```{r}
gt_tbl |>
  opt_interactive()
```

:::

## Examples

Some styling is respected in `opt_interactive()`

```{r}
styled <- gt_tbl |>
  tab_options(
    heading.title.font.weight = "bold",
    stub.background.color = "lightblue", 
    table.border.bottom.style = "dotted",
    column_labels.background.color = "pink",
    table.font.weight = "italic",
    stub.font.weight = "bolder",
    table_body.vlines.color = "brown",
    table_body.vlines.style = "dashed"
  )
```

::: {.panel-tabset}
## Html

```{r}
#| echo: false
styled
```

## Interactive

```{r}
#| echo: false
styled %>% opt_interactive()
```

:::




# Current limitations

-   Some features like `tab_style()` may not be fully supported.

-   `summary_rows()` and `grand_summary_rows()` have yet to be implemented.

-   Your interactive table may be visually different from your non-interactive table.
