---
title: "gt interactive tables"
format: html
description: >
  An overview of interactive tables with gt
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

gt provides an option to make interactive html tables via the [reactable](https://glin.github.io/reactable/index.html) package.

It also provides an interactive to creating gt tables as plots.

Let's use the following base for our gt table.

```{r}
library(gt)
gt_tbl <- exibble |>
  gt()
```

To create an interactive table, you have simply have to pipe `opt_interactive()` to your existing gt pipeline.

```{r}
gt_tbl
```

```{r}
gt_tbl |>
  opt_interactive()
```



# Current limitations

* Some features like `tab_style()` may not be fully supported. 

* `summary_rows()` and `grand_summary_rows()` have yet to be implemented.

* Your interactive table may be visually different from your non-interactive table.


# Visual tests static vs interactive

# `math formulas()`

@tbl-md contains many things.

```{r}
#| tbl-cap: "A table"
#| label: tbl-md
#| code-fold: true

df <- data.frame(
  idx = 1:37,
  l_time_domain = 
    c(
      "$1$",
      "${{\\bf{e}}^{a\\,t}}$",
      "${t^n},\\,\\,\\,\\,\\,n = 1,2,3, \\ldots$",
      "${t^p}, p > -1$",
      "$\\sqrt t$",
      "${t^{n - \\frac{1}{2}}},\\,\\,\\,\\,\\,n = 1,2,3, \\ldots$",
      "$\\sin \\left( {at} \\right)$",
      "$\\cos \\left( {at} \\right)$",
      "$t\\sin \\left( {at} \\right)$",
      "$t\\cos \\left( {at} \\right)$",
      "$\\sin \\left( {at} \\right) - at\\cos \\left( {at} \\right)$",
      "$\\sin \\left( {at} \\right) + at\\cos \\left( {at} \\right)$",
      "$\\cos \\left( {at} \\right) - at\\sin \\left( {at} \\right)$",
      "$\\cos \\left( {at} \\right) + at\\sin \\left( {at} \\right)$",
      "$\\sin \\left( {at + b} \\right)$",
      "$\\cos \\left( {at + b} \\right)$",
      "$\\sinh \\left( {at} \\right)$",
      "$\\cosh \\left( {at} \\right)$",
      "${{\\bf{e}}^{at}}\\sin \\left( {bt} \\right)$",
      "${{\\bf{e}}^{at}}\\cos \\left( {bt} \\right)$",
      "${{\\bf{e}}^{at}}\\sinh \\left( {bt} \\right)$",
      "${{\\bf{e}}^{at}}\\cosh \\left( {bt} \\right)$",
      "${t^n}{{\\bf{e}}^{at}},\\,\\,\\,\\,\\,n = 1,2,3, \\ldots$",
      "$f\\left( {ct} \\right)$",
      "${u_c}\\left( t \\right) = u\\left( {t - c} \\right)$",
      "$\\delta \\left( {t - c} \\right)$",
      "${u_c}\\left( t \\right)f\\left( {t - c} \\right)$",
      "${u_c}\\left( t \\right)g\\left( t \\right)$",
      "${{\\bf{e}}^{ct}}f\\left( t \\right)$",
      "${t^n}f\\left( t \\right),\\,\\,\\,\\,\\,n = 1,2,3, \\ldots$",
      "$\\displaystyle \\frac{1}{t}f\\left( t \\right)$",
      "$\\displaystyle \\int_{{\\,0}}^{{\\,t}}{{\\,f\\left( v \\right)\\,dv}}$",
      "$\\displaystyle \\int_{{\\,0}}^{{\\,t}}{{f\\left( {t - \\tau } \\right)g\\left( \\tau  \\right)\\,d\\tau }}$",
      "$f\\left( {t + T} \\right) = f\\left( t \\right)$",
      "$f'\\left( t \\right)$",
      "$f''\\left( t \\right)$",
      "${f^{\\left( n \\right)}}\\left( t \\right)$"
    ),
  l_laplace_s_domain =
    c(
      "$$\\frac{1}{s}$$",
      "$$\\frac{1}{{s - a}}$$",
      "$$\\frac{{n!}}{{{s^{n + 1}}}}$$",
      "$$\\frac{{\\Gamma \\left( {p + 1} \\right)}}{{{s^{p + 1}}}}$$",
      "$$\\frac{{\\sqrt \\pi }}{{2{s^{\\frac{3}{2}}}}}$$",
      "$$\\frac{{1 \\cdot 3 \\cdot 5 \\cdots \\left( {2n - 1} \\right)\\sqrt \\pi }}{{{2^n}{s^{n + \\frac{1}{2}}}}}$$",
      "$$\\frac{a}{{{s^2} + {a^2}}}$$",
      "$$\\frac{s}{{{s^2} + {a^2}}}$$",
      "$$\\frac{{2as}}{{{{\\left( {{s^2} + {a^2}} \\right)}^2}}}$$",
      "$$\\frac{{{s^2} - {a^2}}}{{{{\\left( {{s^2} + {a^2}} \\right)}^2}}}$$",
      "$$\\frac{{2{a^3}}}{{{{\\left( {{s^2} + {a^2}} \\right)}^2}}}$$",
      "$$\\frac{{2a{s^2}}}{{{{\\left( {{s^2} + {a^2}} \\right)}^2}}}$$",
      "$$\\frac{{s\\left( {{s^2} - {a^2}} \\right)}}{{{{\\left( {{s^2} + {a^2}} \\right)}^2}}}$$",
      "$$\\frac{{s\\left( {{s^2} + 3{a^2}} \\right)}}{{{{\\left( {{s^2} + {a^2}} \\right)}^2}}}$$",
      "$$\\frac{{s\\sin \\left( b \\right) + a\\cos \\left( b \\right)}}{{{s^2} + {a^2}}}$$",
      "$$\\frac{{s\\cos \\left( b \\right) - a\\sin \\left( b \\right)}}{{{s^2} + {a^2}}}$$",
      "$$\\frac{a}{{{s^2} - {a^2}}}$$",
      "$$\\frac{s}{{{s^2} - {a^2}}}$$",
      "$$\\frac{b}{{{{\\left( {s - a} \\right)}^2} + {b^2}}}$$",
      "$$\\frac{{s - a}}{{{{\\left( {s - a} \\right)}^2} + {b^2}}}$$",
      "$$\\frac{b}{{{{\\left( {s - a} \\right)}^2} - {b^2}}}$$",
      "$$\\frac{{s - a}}{{{{\\left( {s - a} \\right)}^2} - {b^2}}}$$",
      "$$\\frac{{n!}}{{{{\\left( {s - a} \\right)}^{n + 1}}}}$$",
      "$$\\frac{1}{c}F\\left( {\\frac{s}{c}} \\right)$$",
      "$$\\frac{{{{\\bf{e}}^{ - cs}}}}{s}$$",
      "${{\\bf{e}}^{ - cs}}$",
      "${{\\bf{e}}^{ - cs}}F\\left( s \\right)$",
      "${{\\bf{e}}^{ - cs}}{\\mathcal{L}}\\left\\{ {g\\left( {t + c} \\right)} \\right\\}$",
      "$F\\left( {s - c} \\right)$",
      "${\\left( { - 1} \\right)^n}{F^{\\left( n \\right)}}\\left( s \\right)$",
      "$\\int_{{\\,s}}^{{\\,\\infty }}{{F\\left( u \\right)\\,du}}$",
      "$\\displaystyle \\frac{{F\\left( s \\right)}}{s}$",
      "$F\\left( s \\right)G\\left( s \\right)$",
      "$\\displaystyle \\frac{{\\displaystyle \\int_{{\\,0}}^{{\\,T}}{{{{\\bf{e}}^{ - st}}f\\left( t \\right)\\,dt}}}}{{1 - {{\\bf{e}}^{ - sT}}}}$",
      "$sF\\left( s \\right) - f\\left( 0 \\right)$",
      "${s^2}F\\left( s \\right) - sf\\left( 0 \\right) - f'\\left( 0 \\right)$",
      "${s^n}F\\left( s \\right) - {s^{n - 1}}f\\left( 0 \\right) - {s^{n - 2}}f'\\left( 0 \\right) \\cdots - s{f^{\\left( {n - 2} \\right)}}\\left( 0 \\right) - {f^{\\left( {n - 1} \\right)}}\\left( 0 \\right)$"
    )
) |>
  gt() |>
  fmt_markdown() |>
  cols_label(
    idx = "",
    l_time_domain = md("$f\\left( t \\right) = {\\mathcal{L}^{\\,\\, - 1}}\\left\\{ {F\\left( s \\right)} \\right\\}$"),
    l_laplace_s_domain = md("$F\\left( s \\right) = \\mathcal{L}\\left\\{ {f\\left( t \\right)} \\right\\}$")
  ) |>
  tab_header(title = "A Table of Laplace Transforms") |>
  tab_source_note(
    source_note = md("The hyperbolic functions: $\\cosh \\left( t \\right) = \\frac{{{{\\bf{e}}^t} + {{\\bf{e}}^{ - t}}}}{2}$ , $\\sinh \\left( t \\right) = \\frac{{{{\\bf{e}}^t} - {{\\bf{e}}^{ - t}}}}{2}$")
  ) |>
  cols_align(align = "center")

```
